A Descriptive Analysis of How Psychology Journals Handle Post-Publication

### 2024-01-05 manuscript/manuscript_NHB.Rmd
manuscript draft
updating changes

---
title: "A Descriptive Analysis of How Psychology Journals Handle Post-Publication Critiques"
author: "Whamond, A., Vazire, S., Clarke, B., Moodie, N., Schiavone, S., Thibault, R. T., & Hardwicke, T. E."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include = FALSE}
#renv::activate()

library(papaja) # article template
library(tidyverse)
library(ggplot2) # figure creation
library(ggrepel) # for ggplot text labels
library(patchwork) # for multipanel plot layouts
library(RColorBrewer) # data visualisation
library(dplyr)
library(kableExtra) # for tables
library(knitr) # for literate programming
library(janitor) # for data munging
library(tidylog) # for inline code feedback
library(here) # for finding files
library(binom) # for calculating CIs
```

```{r echo=FALSE, message=FALSE}
# Load processed data files
#'ProjCode', 'GitClone', 'PROJ_2023_psych-journal-critique',
d_articles <- read_csv(here('data','processed','d_articles.csv'), show_col_types = F) # article data 
d_journals <- read_csv(here('data','processed','d_journals.csv'), show_col_types = F) # journal data 
```

@@

@@

## Journal Policy

For descriptive statistics on journal subdisciplines, post-publication critique acceptance, and COPE membership for both prominent and randomly-selected samples, see Supplemetary Information ##. A full list of the journals included in this study, along with their post-publication critique options and restrictions, is provided in Supplemetary Information ##.

@@

In our sample of 100 prominent journals, the median 2021 Journal Impact Factor was `r promJournalCharacteristics %>% pull(promMedian_jif)` (IQR = `r promJournalCharacteristics %>% pull(promIQR_jif)`, range = `r promJournalCharacteristics %>% pull(promMin_jif)` - `r promJournalCharacteristics %>% pull(promMax_jif)`). As this represents the entire population of interest for this section of the study, confidence intervals (CIs) are not included for this section of the study. Figure *XX* shows theobserved journal policies and, of the explicitly offered critique options identified, how restrictions on length and time-to-submit were stated.

@@

```
As shown in Figure *XX*, post-publication critique policies were explicitly stated by `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "YES") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_yesE_LL, "%, ", ci_yesE_UL, "%]")`), and we inferred that they were implicitly accepted (as denoted by archival examples) in `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "ARCHIVE") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_yesI_LL, "%, ", ci_yesI_UL, "%]")`). `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "NO IMPLICIT") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_noI_LL, "%, ", ci_noI_UL, "%]")`) had no advertised post-publication critique policy, and `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "NO EXPLICIT") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_noE_LL, "%, ", ci_noE_UL, "%]")`) explicitly stated that they did not accept critiques (*Clinical Psychology: Science and Practice and Psychological Bulletin*, also included in sample for prominent journals). `r paste0(random2types, "%")` of journals accepted two types (`r paste0(round(confidence_level*100), "% CI [", ci_2typ_LL, "%, ", ci_2typ_UL, "%]")`); *Cortex* offered commentaries and 'verification reports' (Chambers, 2020) and *Journal of Cognition and Culture* offered commentaries and letters.

@@

## PPC Prevalence (results cont.)

@@

>>>>

---
title: "A Descriptive Analysis of How Psychology Journals Handle Post-Publication
  Critiques"
author: Whamond, A., Vazire, S., Clarke, B., Moodie, N., Schiavone, S., Thibault,
  R. T., & Hardwicke, T. E.
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  word_document: default
abstract: "Fostering critical discussion is essential to upholding the ethos of science
  as an unbiased, collaborative, and self-correcting endeavor. Journals can support
  this by offering options for post-publication critique, such as commentaries or
  letters, allowing peers to voice substantiated concerns of empirical articles. We
  provide a descriptive, cross-sectional analysis of post-publication critique policies
  and prevalence across both prominent randomly-selected and psychology journals.
  We found that 38% of prominent and 23% of randomly-selected psychology journals
  offered explicit options for post-publication critique. Length and time-to-submit
  limits were often not stated. Critique prevalence was rare with only a single event
  identified among two samples of 101 articles. Given the widespread recognition of
  the replication crisis in psychology, allowing avenues for criticisms to be voiced
  may help increase the credibility within the field. To support transparency and
  self-correction, psychology journals should offer clearly stated post-publication
  critique options and encourage their use.\n"
keywords: "post-publication critique, peer review, psychology, meta-research, journal
  policy, self-correction, scientific criticism"

bibliography: references.bib
link-citations: true
csl: nature.csl

floatsintext: false
figurelist: false
tablelist: false
footnotelist: false
linenumbers: true
mask: true
draft: true
always_allow_html: true
urlcolor: blue
---

```{r setup, include = FALSE}

library(papaja) # article template
library(tidyverse)
library(ggplot2) # figure creation
library(ggrepel) # for ggplot text labels
library(patchwork) # for multipanel plot layouts
library(RColorBrewer) # data visualisation
library(dplyr)
library(kableExtra) # for tables
library(knitr) # for literate programming
library(janitor) # for data munging
library(tidylog) # for inline code feedback
library(here) # for finding files
library(binom) # for calculating CIs
library(citation) # for organizing citations
library(tinytex)
```

```{r echo=FALSE, message=FALSE}
# Load processed data files
#'ProjCode', 'GitClone', 'PROJ_2023_psych-journal-critique',
d_articles <- read_csv(here('data', 'processed', 'd_articles.csv'), show_col_types = F) # article data 
d_journals <- read_csv(here('data', 'processed', 'd_journals.csv'), show_col_types = F) # journal data

@@

```{r echo = FALSE, message = FALSE}
# Create Table for describing the different common PPC types and restrictions (based on Hardwicke et al., 2022 findings)
Type <- c("Letters", "Commentaries", "Web Comments")
MedianLength <- c("500", "1200", "600")
MedianTime <- c("8", "26", "NA")
MedianRefs <- c("6", "15", "8")
PeerReview <- c("20%", "55%", "0%")
DOIs <- c("Yes", "Yes", "No")
Utility <- c("Concise criticisms submitted shortly after target article published", "More in-depth critical reviews", "Flexible length with almost immediate visibility to readership")

PPCtypes <- data.frame(Type, MedianLength, MedianTime, MedianRefs, PeerReview, DOIs, Utility)
```

``` {r echo = FALSE, results = 'asis'}
PPCtypes %>%
  kbl(align = c(rep("l", 1), rep("c", 5), rep("l", 1)), caption = "Common post-publication critique types including median restrictions, peer review policies, and descriptions",
      col.names = c("Type", "Length", "Time-to-submit", "References", "Peer-review", "DOI", "Description")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE) %>%
  row_spec(0, bold = TRUE, align = "c")

@@

## Part 1: Post-Publication Critique Policies

For descriptive statistics on journal subdisciplines, post-publication critique acceptance, and COPE membership for both prominent and randomly-selected samples, see Supplementary Information ##. A full list of the journals included in this study, along with their post-publication critique options and restrictions, is provided in Supplementary Information ##.

@@

In our sample of 100 prominent journals, the median 2021 Journal Impact Factor was `r promJournalCharacteristics %>% pull(promMedian_jif)` (IQR = `r promJournalCharacteristics %>% pull(promIQR_jif)`, range = `r promJournalCharacteristics %>% pull(promMin_jif)` - `r promJournalCharacteristics %>% pull(promMax_jif)`). As this represents the entire population of interest for this section of the study, confidence intervals (CIs) are not included for this section of the study. Figure *XX* shows the observed journal policies and, of the explicitly offered critique options identified, how restrictions on length and time-to-submit were stated.

@@

```
As shown in Figure *XX*, post-publication critique policies were explicitly stated by `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "YES") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_yesE_LL, "%, ", ci_yesE_UL, "%]")`), and we inferred that they were implicitly accepted (as denoted by archival examples) in `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "ARCHIVE") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_yesI_LL, "%, ", ci_yesI_UL, "%]")`). `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "NO IMPLICIT") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_noI_LL, "%, ", ci_noI_UL, "%]")`) had no advertised post-publication critique policy, and `r dj_random %>% filter(ppc_type == "A") %>% summarise(percentage = paste0(sum(has_ppc == "NO EXPLICIT") / n() * 100, "%"))` of journals (`r paste0(round(confidence_level*100), "% CI [", ci_noE_LL, "%, ", ci_noE_UL, "%]")`) explicitly stated that they did not accept critiques (*Clinical Psychology: Science and Practice and Psychological Bulletin*, also included in sample for prominent journals). `r paste0(random2types, "%")` of journals accepted two types (`r paste0(round(confidence_level*100), "% CI [", ci_2typ_LL, "%, ", ci_2typ_UL, "%]")`); *Cortex* offered commentaries and 'verification reports' [@Chambers2020] and *Journal of Cognition and Culture* offered commentaries and letters.

@@

## Part 2: Post-Publication Critique Prevalence
```{r echo = FALSE, message=FALSE}
# Filter articles by sample and calculate CIs

# For articles randomly sampled from prominent psychology journals
da_prominent <- d_articles %>%
  filter(exclude == "RETAIN") %>% # remove excluded articles
  filter(sample_id == "prominent") # keep only articles from prominent journals

sample_size_a <- 101

# Linked PPCs
prominent_linked <- da_prominent %>% # counts number of linked PPC events
  filter(ppc_linked == "TRUE") %>%
  count()

ci_linked <- binom.confint(prominent_linked$n, sample_size_a, method = "wilson", conf.level = confidence_level)

# Convert the lower and upper limits to percentages and round them to one decimal place
ci_linked_LL <- round(ci_linked$lower*100, 1)
ci_linked_UL <- round(ci_linked$upper*100, 1)

# Instance PPCs
prominent_instance <- da_prominent %>% # counts number of linked PPC events
  filter(exclude_reason == "IS PPC") %>%
  count()

ci_instance <- binom.confint(prominent_instance$n, sample_size_a, method = "wilson", conf.level = confidence_level)

# Convert the lower and upper limits to percentages and round them to one decimal place
ci_instance_LL <- round(ci_instance$lower*100, 1)
ci_instance_UL <- round(ci_instance$upper*100, 1)

# For articles randomly sampled from all psychology journals
da_random <- d_articles %>%
  filter(exclude == "RETAIN") %>% # remove excluded articles
  filter(sample_id == "random") # keep only randomly-selected articles

# Linked PPCs
random_linked <- da_random %>%
  filter(ppc_linked == "TRUE") %>%
  count()

ci_linkedR <- binom.confint(random_linked$n, sample_size_a, method = "wilson", conf.level = confidence_level)

# Convert the lower and upper limits to percentages and round them to one decimal place
ci_linkedR_LL <- round(ci_linkedR$lower*100, 1)
ci_linkedR_UL <- round(ci_linkedR$upper*100, 1)

# Instance PPCs
random_instance <- da_random %>% # counts number of linked PPC events
  filter(exclude_reason == "IS PPC") %>%
  count()

ci_instanceR <- binom.confint(random_instance$n, sample_size_a, method = "wilson", conf.level = confidence_level)

# Convert the lower and upper limits to percentages and round them to one decimal place
ci_instanceR_LL <- round(ci_instanceR$lower*100, 1)
ci_instanceR_UL <- round(ci_instanceR$upper*100, 1)
```
**2a) Articles Published in Prominent Psychology Journals**

Prevalence was defined in two ways: linked critiques (links to critiques appearing on same webpage as target article) and instances of critiques (articles that are themselves critiques). Only one linked post-publication critique event was found within our sample of 101 articles published in prominent psychology journals, yielding a prevalence estimate of `r da_prominent %>% summarise(percentage = paste0(round(sum(as.integer(ppc_linked == "TRUE")) / n() * 100, 1), "%"))` (`r paste0(round(confidence_level*100), "% CI [", ci_linked_LL, "%, ", ci_linked_UL, "%]")`). No instances of post-publication critiques were found (0%, `r paste0(round(confidence_level*100), "% CI [", ci_instance_LL, "%, ", ci_instance_UL, "%]")`).

The single linked critique found appeared under the 'Recommended articles' sidebar of an article published in the Journal of Environmental Psychology webpage (for a depiction of how the article was linked, see Appendix F). This example was published as a letter[@Berto2020] and addressed concerns over incorrect citation of the critiquing author’s previous work and methodology used in the target article [@Hicks2020].

**2b) Articles Published in Any Psychology Journal**

No events of either linked or instances of post-publication critiques were identified in our random sample of 101 articles, resulting in a prevalence estimate of 0% (`r paste0(round(confidence_level*100), "% CI [", ci_instanceR_LL, "%, ", ci_instanceR_UL, "%]")`) for both classifications.

@@

# References
